A generalization of the Bradley-Terry model for draws in chess

### 2020-10-04 inst/jebo_chess_rebuttal.txt

But frankly I don't want to open the can of worms that would be
comparison between Elo and Bradley-Terry.  I had a protracted argument
on this issue with a chess enthusiast; let us say the discussion was
intense and robust.  The problem is that Elo was designed specifically
to downweight old data and this downweighting is what users of such
metrics want.  Unfortunately, what my correspondent wanted from
Bradley-Terry strengths was pretty much Elo with all its
arbitrariness: a somewhat circular position.

> 
> Precise estimation will hinge on a large number of games, but this
> will be in tension with the assumption of time-invariant
> player-strengths.  The author should explore this tension.
>

I am in total agreement with this observation.  My thinking on this
issue has been strongly affected by my recent work applying
Bradley-Terry to Formula 1 motor racing [I have a manuscript under
review at JQAS].  In F1, drivers generally start their careers as weak
competitors, become stronger with experience, and finally become
uncompetitive due to age [although frankly Lewis Hamilton at 35yo
seems to buck this general trend, as recent events in Tuscany would
indicate].  I would suggest that chess players have a similar career
arc, although Formula 1 Teams typically take a dim view of
underperforming drivers---whom they rapidly and unceremonially drop.
The career arc for these competitors would have a truncated right hand
side, unlike chess players who typically decline gradually over
decades [although I would observe that Kasparov's current FIDE rating
of 2812 is not that much less than his peak of 2851].

Specifically, the dataset in question stretches from 1981 to 2017, a
very wide range.  Generalized Bradley-Terry, together with reified
entities, is capable of handling strengths varying over a career: one
would introduce a "mid-career monster", or perhaps an "experience
monster".  I would add here that monsters' strengths are inherently
positive, so one could deal with waning strength with age by
introducing a "youth monster" who helps younger players.  But such
analysis is, I would suggest, firmly in the category of further work
(although the revision does address a time-dependent component in the
form of a "rest monster").  I have added a discussion of the
identification issue to the manuscript (RED), in two parts: firstly
when introducing the dataset, and secondly under Further Work.

@@

> 2) The author neglects to motivate the paper beyond its application
> within the chess world.  The tournament literature's applications lie
> primarily, though definitely not exclusively, within compensation and
> labor.  How could this extension be used within that context?  Where
> might economists see the first-mover advantage in action? Where might
> the extension to allow draws be especially lucrative? Etc.
>

(manuscript changes addressing this comment are marked in ORANGE).

The submission discusses applications to MasterChef Australia,
consumer behaviour and mating preference, but I was not aware of
tournament literature as a distinct academic field.  I have been
studying Lazear and Rosen, and Szymanksi 2003, and pondering how
reified monsters could help understanding tournaments in a workplace
remuneration context.  Lazear and Rosen supply a good rationale to
consider only rank in situations such as Formula 1 motor racing.

### 2020-10-04 inst/jebo_chess_revision2.tex
more accounting for referee comments

\begin{abstract}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work that uses additional information such as
historical results including number of moves.
\end{abstract}

@@

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).  \textcolor{orange}{First-mover advantage is closely
  related to incumbent advantage, and the techniques presented here
  have been applied to ecological incumbency~\cite{hankin2010}.}

@@

\citet{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$\draw=\nu\sqrt{p_ip_j}$.  However, \citet{joe1990} uses likelihood
methods to show that this model does not fit data well.  Here we note
that such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.

@@

\begin{table}
\textcolor{DarkGreen}{
  \caption{Results for won games in Cura\c{c}ao 1962  between
      rested and unrested players    \label{restmonsterfisher}}
  \centering
  \fbox{%
    \begin{tabular}{l|cc}\\
      & White rested & Black rested \\ \hline
      White wins  & $9$   & $ 3$   \\
      Black wins  & $0$   & $13$   \\
  \end{tabular}}
  }
\end{table}

@@

\cite{moul2009} present detailed Monte Carlo simulations to assess the
likely impact of a putative Soviet cartel on tournament outcomes.  In
principle, the present system of likelihood and reified entities would
admit such simulations and it might be interesting to compare results.
\textcolor{blue}{\cite{moul2009} go on to compare round robin
  tournaments to knockout format events, searching for evidence of a
  Soviet draw cartel.  By only considering international tournaments
  [where Soviets have incentive to play differently against fellow
    Soviets or Westerners] the analysis presented here does not allow
  me to do this but the technique could easily be applied to a more
  informative dataset.}

@@

\section*{References}

>>>>

\begin{abstract}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work.  \textcolor{blue}{I also present
  scoreline evidence for the effectiveness of such a drawing cartel:
  noncollusive games are less detrimental to future play than
  collusive games ($p<10^{-6}$).}
\end{abstract}

@@

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage,
winning about 37.5\% of tournament games (black wins 27.6\%); and
secondly, many chess games, about 39.4\%, are drawn (figures from
Chessbase).  \textcolor{orange}{First-mover advantage is closely
  related to incumbent advantage, and the techniques presented here
  have been applied to ecological incumbency~\cite{hankin2010}.  More
  generally, \cite{szymanski2003} considers sporting contests as
  examples of competitive situations in which a contribution of some
  kind [effort, investment, attendance of the elite, etc] is elicited
  from contestants who may as a result win a prize.
  \citeauthor{szymanski2003} points out that this paradigm arguably
  covers executive corporate promotion, interviews for a position in
  an office, and most sporting tournaments.  As such, the phenomenon
  of collusion and indeed drawing in pairwise comparison is surely an
  interesting phenomenon.}

@@

\citet{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$\draw=\nu\sqrt{p_ip_j}$.  However, \citet{joe1990} uses likelihood
methods to show that this model does not fit data well.  Here we note
that such a model automatically gives stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.

@@

\begin{table}
\textcolor{DarkGreen}{
  \caption{Results for the 25 won games (out of 49 played) in
    Cura\c{c}ao 1962 between rested and unrested
    players \label{restmonsterfisher}} \centering \fbox{%
    \begin{tabular}{l|cc}\\
      & White rested & Black rested \\ \hline
      White wins  & $9$   & $ 3$   \\
      Black wins  & $0$   & $13$   \\
  \end{tabular}}
  }
\end{table}

@@

\cite{moul2009} present detailed Monte Carlo simulations to assess the
likely impact of a putative Soviet cartel on tournament outcomes.  In
principle, the present system of likelihood and reified entities would
admit such simulations and it might be interesting to compare results.
\textcolor{blue}{\cite{moul2009} go on to compare round robin
  tournaments to knockout format events, searching for evidence of a
  Soviet draw cartel.  The analysis presented here, only considering
  international tournaments [where Soviets have incentive to play
    differently against fellow Soviets or Westerners] does not admit
  this phenomenon, but the technique could easily be applied to a more
  informative dataset.}

@@
